---
title: "Code from: Genetic and environmental factors shape rates of plasticity: the temporal dynamics of opsin gene expression in aquatic environments"
author: "César Bertinetti"
date: "March 2025"
output:
  html_document:
    df_print: paged
  pdf_document:
    latex_engine: xelatex
---

<style type="text/css">

h1.title {
  font-size: 28px;
  color: Black;
  text-align: justify;
}
h4.author { /* Header 4 - and the author and data headers use this too  */
    font-size: 18px;
  font-family: "Helvetica", Times, serif;
  color: Black;
  text-align: justify;
}
h4.date { /* Header 4 - and the author and data headers use this too  */
  font-size: 18px;
  font-family: "Helvetica", Times, serif;
  color: Black;
  text-align: justify;
}
</style>

<style>
pre {
  font-size: 14px;
}
</style>

<br/>

The datasets needed to reproduce this analysis are deposited on Bertinetti et al. 2025. Data from: *Genetic and environmental factors shape rates of plasticity: the temporal dynamics of opsin gene expression in aquatic environments* [Dataset]. Zenodo.

Following packages are required:

```{r include=FALSE}
pak <- c("readr", "stringr", "matrixStats","dplyr","tidyr","plotrix","RColorBrewer","MASS",
         "MESS","zoo","car","Hmisc","corrplot","agricolae","nlme","ggplot2","gridExtra",
         "stringr","tidyr", "pracma","visreg", "nlme", "sjstats", "performance", "lme4","boot",
         "rsq","partR2","grid","effectsize","devtools", "readxl","broom","tidyverse","broom",
         "ggfortify", "vegan","ggpubr","ggdist","colorspace","ragg","ggtext","nls2","minpack.lm")
lapply(pak, require, character.only = TRUE) #check if packages are installed
``` 

```{r, include=TRUE}

R.version$platform
R.version$version.string
installed.packages()[names(sessionInfo()$otherPkgs), "Version"]

```

## Read Opsin Gene Expression and Plot Distribution ##

The "Opsin gene expression.csv" contains the raw opsin gene expression values for all individuals

```{r}
## READ MASTERTABLE GENERATED FROM COMBINE qPCR
opsin <- read.table("Opsin gene expression.csv", header=T, sep =",", dec=".",stringsAsFactors = T) [,-1]


# Combine controls - this part can be left out for further resolutoin before 20 dph
for (i in 1:nrow(opsin)) {
  if(opsin$Light[i] == "WR" & opsin$Age[i] < 21){
    opsin$Light[i] <- "WW"
  }else if (opsin$Light[i] == "RW" & opsin$Age[i] < 21){
    opsin$Light[i] <- "RR"
  } else {
    opsin$Light[i] <- opsin$Light[i] 
  }
}

## SPLIT TO CREATE "BROAD GROUPS" THAT HELP WHEN PlOTTING (ALL TREATMENTS)
opsin <- opsin %>%
  unite("onto", c("Stage","Age"), sep="", remove=F) %>%
  unite("combi", c("Light", "onto","Spec"), sep="-", remove=F)
opsin <- opsin[,-5] # removes onto column

## ADD DAYS AFTER CHANGE AS A VARIABLE
for (i in 1:nrow(opsin)) {
  if(opsin$Stage[i] == "E" & opsin$Age[i] >= 20){
    opsin$day[i] <- as.numeric(opsin$Age[i] - 20)
  }else if (opsin$Stage[i] == "L"){
    opsin$day[i] <- as.numeric(opsin$Age[i] - 35)
  } else {
    opsin$day[i] <- NA
  }
}

opsin1 <- opsin[,c(3,8:18)] ### EXTRACT IMPORTANT VARIABLES BEFORE AGGREGGATING

## FIRST MAKES SENSE TO AGGREGATE REPLICATES

## THIS IS A SIMPLE FUNCTION TO GET M + SE + SAMPLE SIZE
momred = function(x) {
  c(m = mean(x), n = length(x), se = std.error(x))
}

## HERE ONLY FOR SC NOW
mean_exp <- aggregate(cbind(SC_sws1,SC_sws2b,SC_sws2a,PSI_SC) ~ combi, data = opsin1, momred)
mean_exp <- do.call(data.frame, mean_exp)

opsin_mean <- unique(merge( opsin[,c(3:7,19)],mean_exp, by="combi")) ## MERGE TO RETTACH METADATA

opsin_mean <- opsin_mean %>% ## SPLIT TO CREATE "BROAD GROUPS" THAT HELP WHEN PlOTTING (ALL TREATMENTS)
  unite("brgroups", c("Light", "Stage","Spec"), sep="-", remove=F)


```

```{r, warning=F}
## SUBSETTIN FOR ONLY EARLY TREATMENT
sp <- "XIL" # "NIC"
st <- "E"

#opsin_early <- subset(opsin_mean, Stage == st )
opsin_early <- subset(opsin_mean, Spec == sp & Stage == st & Age < 51)
#genes <- c("sws1","sws2b","sws2a","rh2b","rh2a","lws")
genes <- c("SC_sws1","SC_sws2b","SC_sws2a") ## SAME BUT PROP OF SC OR DC
P_plots <- list()

for (i in 1:length(genes)) {
  
  b <- as.name(paste(as.character(genes[i]),".m", sep="")) ## creates column name for variable "mean"
  c <- as.name(paste(as.character(genes[i]),".se", sep=""))
  if (i==7) {
    pl  <- ggplot(opsin_early, aes(x=Age, y=eval(as.name(b)), group=brgroups, color=brgroups)) +
      geom_line()+
      theme_classic()+
      geom_pointrange(aes(ymin=eval(as.name(b))-eval(as.name(c)), ymax=eval(as.name(b))+eval(as.name(c))))+
      ylab(as.name(b))+
      scale_x_continuous(breaks=seq(0,50,5)) +
      geom_vline(xintercept=c(20,35), linetype=2) +
      theme(legend.position = "inside",legend.position.inside=c(0.8,0.8)) +
      ylim(0,1)
    P_plots[[i]] <- ggplotGrob (pl) ## save to be able to plot outside of loop
  } else {
    pl  <- ggplot(opsin_early, aes(x=Age, y=eval(as.name(b)), group=brgroups, color=brgroups)) +
      geom_line()+
      theme_classic()+
      geom_pointrange(aes(ymin=eval(as.name(b))-eval(as.name(c)), ymax=eval(as.name(b))+eval(as.name(c))))+
      ylab(as.name(b))+
      scale_x_continuous(breaks=seq(0,50,5)) +
      geom_vline(xintercept=c(20,35), linetype=2) +
      ylim(0,1)+
      theme(legend.position = "none") ## don't want legend in all plots
    P_plots[[i]] <- ggplotGrob (pl)
  }
  #  P_plots[[i]] <- ggplotGrob (pl)
  #print(P_plots[[i]])
}
```

```{r}
grid.arrange(grobs=P_plots, ncol=3, top = textGrob(sp)) 
```


## Read and visualize single cone opsin predicted sensitivity index ##

```{r, warning=F}

P_plots <- list()

opsin_m <- subset(opsin_mean, Spec == "XIL" & Age < 51)

xil  <- ggplot(opsin_m, aes(x=Age, y=PSI_SC.m, group=Light, color=Light)) +
    geom_line()+
    theme_classic()+
    #geom_pointrange(aes(ymin=eval(as.name(b))-eval(as.name(c)), ymax=eval(as.name(b))+eval(as.name(c))))+
    geom_pointrange(aes(ymin=PSI_SC.m-PSI_SC.se, ymax=PSI_SC.m+PSI_SC.se))+
    labs(x = " Age (days post-hatch)", y = "Single Cone Predicted Sensitivity Index (nm)", title="Xiloá (Clear Lake)")+
    scale_x_continuous(breaks=seq(0,50,5)) +
    geom_vline(xintercept=c(20,35), linetype=2) +
    theme(legend.position = "none", text= element_text(size=12)) +
    ylim(350,500)
P_plots[[1]] <- ggplotGrob (xil) ## save to be able to plot outside of loop

opsin_m <- subset(opsin_mean, Spec == "NIC")

nic  <- ggplot(opsin_m, aes(x=Age, y=PSI_SC.m, group=Light, color=Light)) +
  geom_line()+
  theme_classic()+
  #geom_pointrange(aes(ymin=eval(as.name(b))-eval(as.name(c)), ymax=eval(as.name(b))+eval(as.name(c))))+
  geom_pointrange(aes(ymin=PSI_SC.m-PSI_SC.se, ymax=PSI_SC.m+PSI_SC.se))+
  labs(x = " Age (dph)", y = "Single Cone Predicted Sensitivity Index (nm)", title="Nicaragua (Turbid Lake)")+
  scale_x_continuous(breaks=seq(0,50,5)) +
  geom_vline(xintercept=c(20,35), linetype=2) +
  theme(text= element_text(size=12)) +
  ylim(350,500)
P_plots[[2]] <- ggplotGrob (nic)

```

```{r}

grid.arrange(grobs=P_plots, ncol=2) ## put all panels together

```

## Model mean cone opsin predicted sensitivity index PSI_SC ##


The "LightData.zip" contains the raw irradiance measurements, the normalized irradiance and the photic factors

```{r}
df <- read.csv("PlasticRate-PSI_SC.csv", sep = ",", dec = ".")
### Rename variables
df$y <- df$PSI_SC.m
df$t <- df$day
df$lake <-df$Spec
df$age <-df$Stage
df$light <-df$Light

###  model_key <- nls(y ~ yf + (y0 - yf) * exp(-lambda * t),
###  model_used <- nls(y ~ yf + (Dy0yf) * exp(-lambda * t),

df <- df %>% filter(lake %in% c("NIC", "XIL"))
df <- df %>% filter(light %in% c("RW","WR"))
#df$lake_indicator <- ifelse(df$lake == "nicaragua", 0, 1)
df$lake_indicator <- ifelse(df$lake == "NIC", 0, 1)
df$age_indicator <- ifelse(df$Stage == "E", 0, 1)
df$change_indicator <- ifelse(df$light == "WR", 0, 1)

```


```{r}

full_model <- nlsLM(y ~ (yf + yf_lake * lake_indicator + yf_age * age_indicator + yf_change * change_indicator + 
                        yf_chXlake * change_indicator * lake_indicator + yf_chXage * change_indicator * age_indicator + 
                        yf_lakeXage * lake_indicator * age_indicator + yf_lakeXageXch * change_indicator * lake_indicator * age_indicator) +  
                    (Dy0yf + Dy0yf_lake * lake_indicator + Dy0yf_age * age_indicator + Dy0yf_change * change_indicator + 
                        Dy0yf_chXlake * change_indicator * lake_indicator + Dy0yf_chXage * change_indicator * age_indicator + 
                        Dy0yf_lakeXage * lake_indicator * age_indicator + Dy0yf_lakeXageXch * change_indicator * lake_indicator * age_indicator) * 
                    exp(-(lambda + lambda_lake * lake_indicator + lambda_age * age_indicator + lambda_change * change_indicator + 
                        lambda_CHxLa * change_indicator * lake_indicator + lambda_chXage * change_indicator * age_indicator + 
                        lambda_lakeXage * lake_indicator * age_indicator + lambda_lakeXageXch * change_indicator * lake_indicator * age_indicator) * t),
                  data = df,
                  start = list(Dy0yf = -75, Dy0yf_lake = -20, Dy0yf_age = -34.88, Dy0yf_change = 100, Dy0yf_chXlake = 30, 
                               Dy0yf_chXage = 0, Dy0yf_lakeXage = 0, Dy0yf_lakeXageXch = 0, yf = 465.72, 
                               yf_lake = 1.22, yf_age = 1.22, yf_change = 1.22, yf_chXlake = 1.22, yf_chXage = 1.22, 
                               yf_lakeXage = 1.22, yf_lakeXageXch = 1.22, lambda = 0.0015, lambda_lake = 0.00015, 
                               lambda_age = 0.00015, lambda_change = 0.00015, lambda_CHxLa = 0.00015, lambda_chXage = 0.00015, 
                               lambda_lakeXage = 0.00015, lambda_lakeXageXch = 0.00015),
                  control=nls.lm.control(maxiter = 200, factor = 10, maxfev = 10000, nprint=F))

summary(full_model)

## NO yf_lakeXageXch NO Dy0yf_lakeXageXch
afm_v3 <- nlsLM(y ~ (yf + yf_lake * lake_indicator + yf_age * age_indicator + yf_change * change_indicator + yf_chXlake * change_indicator* lake_indicator+ yf_chXage * change_indicator* age_indicator + yf_lakeXage * lake_indicator* age_indicator  ) +  
                             (Dy0yf + Dy0yf_lake * lake_indicator + Dy0yf_age * age_indicator+ Dy0yf_change * change_indicator + Dy0yf_chXlake * change_indicator* lake_indicator + Dy0yf_chXage * change_indicator* age_indicator + Dy0yf_lakeXage * lake_indicator* age_indicator) * 
                             exp(-(lambda + lambda_lake * lake_indicator + lambda_age * age_indicator+ lambda_change * change_indicator + lambda_CHxLa * change_indicator* lake_indicator + lambda_chXage * change_indicator* age_indicator + lambda_lakeXage * lake_indicator* age_indicator + lambda_lakeXageXch * change_indicator * lake_indicator* age_indicator) * t),
                           data = df,
                           start = list(Dy0yf = -75, Dy0yf_lake = -20, Dy0yf_age = -30, Dy0yf_change = 120, Dy0yf_chXlake = 60, Dy0yf_chXage=0, Dy0yf_lakeXage=0, yf =460, yf_lake = 0, yf_age = 0 ,yf_change = -50, yf_chXlake = -50, yf_chXage = 30, yf_lakeXage=10, 
                                        lambda = 0.8, lambda_lake = 0, lambda_age = -0.7, lambda_change = -0.4, lambda_CHxLa = -.2, lambda_chXage=1, lambda_lakeXage=0, lambda_lakeXageXch=0),
                           control=nls.lm.control(maxiter = 100))

summary(afm_v3)

# noDy0yf_lakeXageXch # noDy0yf_lakeXage  #noyf_lakeXage
afm_v4 <- nlsLM(y ~ (yf + yf_lake * lake_indicator + yf_age * age_indicator + yf_change * change_indicator + yf_chXlake * change_indicator* lake_indicator+ yf_chXage * change_indicator* age_indicator) +  
                                   (Dy0yf + Dy0yf_lake * lake_indicator + Dy0yf_age * age_indicator+ Dy0yf_change * change_indicator + Dy0yf_chXlake * change_indicator* lake_indicator + Dy0yf_chXage * change_indicator* age_indicator) * 
                                   exp(-(lambda + lambda_lake * lake_indicator + lambda_age * age_indicator+ lambda_change * change_indicator + lambda_CHxLa * change_indicator* lake_indicator + lambda_chXage * change_indicator* age_indicator + lambda_lakeXage * lake_indicator* age_indicator + lambda_lakeXageXch * change_indicator * lake_indicator* age_indicator) * t),
                                 data = df,
                                 start = list(Dy0yf = -75, Dy0yf_lake = -20, Dy0yf_age = -30, Dy0yf_change = 120, Dy0yf_chXlake = 60, Dy0yf_chXage=0, yf =460, yf_lake = 0, yf_age = 0 ,yf_change = -50, yf_chXlake = -50, yf_chXage = 30, 
                                              lambda = 0.8, lambda_lake = 0, lambda_age = -0.7, lambda_change = -0.4, lambda_CHxLa = -.2, lambda_chXage=1, lambda_lakeXage=0, lambda_lakeXageXch=0),
                                 control=nls.lm.control(maxiter = 100))

summary(afm_v4)

# noDy0yf_lakeXageXch # noDy0yf_lakeXage  #noyf_lakeXage
# no yf_chXage # yf_chXlake
afm_v5 <- nlsLM(y ~ (yf + yf_lake * lake_indicator + yf_age * age_indicator + yf_change * change_indicator) +  
                  (Dy0yf + Dy0yf_lake * lake_indicator + Dy0yf_age * age_indicator+ Dy0yf_change * change_indicator + Dy0yf_chXlake * change_indicator* lake_indicator + Dy0yf_chXage * change_indicator* age_indicator) * 
                  exp(-(lambda + lambda_lake * lake_indicator + lambda_age * age_indicator+ lambda_change * change_indicator + lambda_CHxLa * change_indicator* lake_indicator + lambda_chXage * change_indicator* age_indicator + lambda_lakeXage * lake_indicator* age_indicator + lambda_lakeXageXch * change_indicator * lake_indicator* age_indicator) * t),
                data = df,
                start = list(Dy0yf = -75, Dy0yf_lake = -20, Dy0yf_age = -30, Dy0yf_change = 120, Dy0yf_chXlake = 60, Dy0yf_chXage=0, yf =460, yf_lake = 0, yf_age = 0 ,yf_change = -50, 
                             lambda = 0.8, lambda_lake = 0, lambda_age = -0.7, lambda_change = -0.4, lambda_CHxLa = -.2, lambda_chXage=1, lambda_lakeXage=0, lambda_lakeXageXch=0),
                control=nls.lm.control(maxiter = 100))

summary(afm_v5)

# noDy0yf_lakeXageXch # noDy0yf_lakeXage  #noyf_lakeXage
# no yf_chXage # yf_chXlake # noyf_lakeXageXch
# no yf_lake yf_age
afm_v6 <- nlsLM(y ~ (yf + yf_change * change_indicator) +  
                  (Dy0yf + Dy0yf_lake * lake_indicator + Dy0yf_age * age_indicator+ Dy0yf_change * change_indicator + Dy0yf_chXlake * change_indicator* lake_indicator + Dy0yf_chXage * change_indicator* age_indicator) * 
                  exp(-(lambda + lambda_lake * lake_indicator + lambda_age * age_indicator+ lambda_change * change_indicator + lambda_CHxLa * change_indicator* lake_indicator + lambda_chXage * change_indicator* age_indicator + lambda_lakeXage * lake_indicator* age_indicator + lambda_lakeXageXch * change_indicator * lake_indicator* age_indicator) * t),
                data = df,
                start = list(Dy0yf = -75, Dy0yf_lake = -20, Dy0yf_age = -30, Dy0yf_change = 120, Dy0yf_chXlake = 60, Dy0yf_chXage=0, yf =460,yf_change = -50, 
                             lambda = 0.8, lambda_lake = 0, lambda_age = -0.7, lambda_change = -0.4, lambda_CHxLa = -.2, lambda_chXage=1, lambda_lakeXage=0, lambda_lakeXageXch=0),
                control=nls.lm.control(maxiter = 100))

summary(afm_v6)

full_additive_model <- nlsLM(y ~ (yf + yf_lake * lake_indicator + yf_age * age_indicator ) +  
                             (Dy0yf + Dy0yf_lake * lake_indicator + Dy0yf_age * age_indicator ) * 
                             exp(-(lambda + lambda_lake * lake_indicator + lambda_age * age_indicator) * t),
                           data = df,
                           start = list(Dy0yf = -95, Dy0yf_lake = -34, Dy0yf_age = -34, 
                                        yf =460, yf_lake = 1, yf_age = 1 ,
                                        lambda = 0.001, lambda_lake = 0, lambda_age = 0.001),
                           control=nls.lm.control(maxiter = 100))

summary(full_additive_model)

null_model <- nlsLM(y ~ (yf ) +  
                             (Dy0yf ) * 
                             exp(-(lambda ) * t),
                           data = df,
                           start = list(Dy0yf = -75, 
                                        yf =460, 
                                        lambda = 0.8))
summary(null_model)

```
## MODEL COMPARISON ## 

```{r}
# Function to compute AIC, BIC, and Degrees of Freedom for nlsLM models
compute_AIC_BIC <- function(model) {
  logLik_val <- logLik(model)  # Log-likelihood
  k <- length(coef(model))  # Number of estimated parameters (degrees of freedom)
  
  AIC_val <- AIC(model)
  BIC_val <- BIC(model)
  
  return(data.frame(Model = deparse(substitute(model)), AIC = AIC_val, BIC = BIC_val, DF = k))
}

```

```{r}
# Function to compute AIC, BIC, and Degrees of Freedom for nlsLM models
compute_AIC_BIC <- function(model) {
  logLik_val <- logLik(model)  # Log-likelihood
  k <- length(coef(model))  # Number of estimated parameters (degrees of freedom)
  
  AIC_val <- AIC(model)
  BIC_val <- BIC(model)
  
  return(data.frame(Model = deparse(substitute(model)), AIC = AIC_val, BIC = BIC_val, DF = k))
}

# Compare all models
aic_bic_results <- rbind(
  compute_AIC_BIC(full_model),
  compute_AIC_BIC(afm_v6),
  compute_AIC_BIC(afm_v3),
  compute_AIC_BIC(afm_v4),
  compute_AIC_BIC(afm_v5),
  compute_AIC_BIC(full_additive_model),
  compute_AIC_BIC(null_model)
)

# Sort by AIC (best model has the lowest AIC)
aic_bic_results <- aic_bic_results[order(aic_bic_results$AIC), ]

# Display results
print(aic_bic_results)

```

```{r}
### BEST MODEL
bmod <- afm_v6
# Compute residual sum of squares (RSS)
rss <- sum(residuals(bmod)^2)

# Compute total sum of squares (TSS)
tss <- sum((df$y - mean(df$y))^2)

# Compute pseudo R²
pseudo_R2 <- 1 - (rss/tss)
pseudo_R2  # 0.96

# Compute RMSE
rmse <- sqrt(mean(residuals(bmod)^2))
rmse  # 6.27
```


## DIAGNOSTICS PLOT FOR MODEL ## 

```{r}
par(mfrow=c(2,2))
# Residuals vs. Fitted
plot(fitted(bmod), residuals(bmod),
     main="Residuals vs. Fitted", xlab="Fitted values", ylab="Residuals")

# QQ-plot for normality
qqnorm(residuals(bmod))
qqline(residuals(bmod))

# Histogram of residuals
hist(residuals(bmod), main="Histogram of Residuals", breaks=20)

# Residuals vs. time (if `t` is a predictor)
plot(df$t, residuals(bmod), main="Residuals over Time",
     xlab="Time", ylab="Residuals")

```

## EXTRACT MODEL ESTIMATES AND CALCULATE STANDARD ERRORS ##

```{r}
# Extract estimates and standard errors
estimates <- summary(bmod)$coefficients[, "Estimate"]
std_errors <- summary(bmod)$coefficients[, "Std. Error"]
# Create a data frame for plotting
params <- data.frame(
  Parameter = names(estimates),
  Estimate = estimates,
  Std_Error = std_errors
)


# Variance-covariance matrix
vcov_matrix <- vcov(bmod)
# Calculate the combined standard error for the sum of two parameters (e.g., lambda_lake & lambda_age)
Nic_late_se <- sqrt(vcov_matrix["lambda", "lambda"] + 
                      vcov_matrix["lambda_age", "lambda_age"] + 
                      2 * vcov_matrix["lambda", "lambda_age"])
Xiloa_early_se <- sqrt(vcov_matrix["lambda", "lambda"] + 
                      vcov_matrix["lambda_lake", "lambda_lake"] + 
                      2 * vcov_matrix["lambda", "lambda_lake"])
xiloa_late_se <- sqrt(
  vcov_matrix["lambda", "lambda"] +
    vcov_matrix["lambda_lake", "lambda_lake"] + 
    vcov_matrix["lambda_age", "lambda_age"] + 
    vcov_matrix["lambda_lakeXage", "lambda_lakeXage"] +
    2 * vcov_matrix["lambda_lake", "lambda_age"] +
    2 * vcov_matrix["lambda", "lambda_age"] +
    2 * vcov_matrix["lambda_lake", "lambda"] +
    2 * vcov_matrix["lambda", "lambda_lakeXage"] +
    2 * vcov_matrix["lambda_lake", "lambda_lakeXage"] +
    2 * vcov_matrix["lambda_age", "lambda_lakeXage"]
)
```


## PLOT MODEL ESTIMATES ##

```{r}
## READ PARAMETERS GENERATED FROM MODEL
plasticity <- read.table("Plasticity_Estimates_Full_Model.csv", header=T, sep =",", dec=".",stringsAsFactors = T) [,-1]

co <- palette.colors()

# Convert categorical variables to factors with defined levels
plasticity$lake <- factor(plasticity$lake, levels = c("Nic", "Xil"))
plasticity$age <- factor(plasticity$age, levels = c("E", "L"))
plasticity$change <- factor(plasticity$change, levels = c("WR", "RW"))
plasticity$Pl_cap <- abs(plasticity$Pl_cap)
#### Plot
# Define color mapping for lakes
#lake_colors <- c("Nic" = "red", "Xil" = "blue")
# change_linetypes <- c("RW" = "solid", "WR" = "dashed")
age_alpha <- c("E" = 1, "L" = 0.5)
lake_colors <- c("Nic" = co[7], "Xil" = co[6])
change_linetypes <- c("WR" = "solid", "RW" = "solid")

# Create a new variable to simplify x-axis labels
#plasticity$x_labels <- ifelse(plasticity$age == "E", "Early", "Late")
plasticity$x_labels <- factor(ifelse(plasticity$change == "WR", "White-to-red", "Red-to-white"), 
                              levels = c( "White-to-red", "Red-to-white"))
# Ensure data is sorted correctly for line plotting
#plasticity <- plasticity[order(plasticity$lake, plasticity$x_labels, plasticity$change), ]

# Create subsets for WR and RW
plasticity_WR <- subset(plasticity, change == "WR")
plasticity_RW <- subset(plasticity, change == "RW")

plasticity_E <- subset(plasticity, age == "E")
plasticity_L <- subset(plasticity, age == "L")
plasticity_L[1,2] <- plasticity_L[1,2] + 0.615

# Define position dodge for x-offset
pos_dodge <- position_dodge(width = 0.2)

# PRATE E (Panel A)
ggplot(plasticity_E, aes(x = x_labels, y = Pl_rate, color = lake, linetype = change)) +
  geom_line(aes(group = lake), linewidth = 1.2, position = pos_dodge) +  # Connect Nic & Xil
  geom_point(aes(alpha = age), size = 3, position = pos_dodge) +  # Apply jitter-like effect
  geom_errorbar(aes(ymin = Pl_rate - Pl_rate_se, ymax = Pl_rate + Pl_rate_se), width = 0.2, position = pos_dodge) +
  scale_color_manual(values = lake_colors) +
  scale_linetype_manual(values = change_linetypes) +
  scale_alpha_manual(values = age_alpha) +
  labs(x = "", y = expression("Plasticity Rate (" * lambda * ")"), title = "") +
  coord_cartesian(ylim = c(0, 2.5)) +
  theme_classic()
```






